A formalization of semi-simplicial and semi-cubical sets in dependent type theory

### 2023-01-17 paper/paper.tex
paper: add \booltype

% The unit type
\newcommand{\unittype}{\ensuremath{\mathsf{unit}}}
\newcommand{\unitpoint}{\ensuremath{\ast}}

@@

The outline of the paper is as follows. First, we briefly describe a free-monoid-style (?) in the fibered presentation of semi-simplicial and semi-cubical sets, and our generalization to $\nu$-sets. We then proceed to explain our indexed construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning with sets inside type theory. We end with some of the finer details of our mechanization.

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the straightforward generalization of augmented semi-simplicial sets and semi-cubical sets to an arbitrary cardinal, where morphisms are sequences of elements of an arbitrary set $\nu$, extended with $\kstar$. That is:

@@

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Coq~\cite{coq22} or Lean~\cite{lean15}. Type theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path and substitutivity of equality as transport. This model supports the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Typical types in type theory are the types of dependent pairs, called $\Sigma$-types, with projections $\hd$ and $\tl$, the types of dependent functions, called $\Pi$-types, a distinguished singleton type called $\textsf{unit}$ type, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers.

@@

\begin{align*}
  \Dom & : \Type_m                                        \\
  \UIP & : \forall x y : \Dom, \forall p q : x = y, p = q
\end{align*}

@@

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:
    \begin{align*}
      \textsf{eq\_refl}        & : \forall x, x = x                                                    \\
      \textsf{UIP\_refl\_unit} & : \forall u : \textsf{unit}, \forall x : u = u, x = \textsf{eq\_refl} \\
      \textsf{UIP\_refl\_bool} & : \forall b : \textsf{bool}, \forall x : b = b, x = \textsf{eq\_refl}
    \end{align*}
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

\begin{align*}
  \textsf{UIP\_refl} & : \forall x:\Dom, \forall p : x = x, p = \textsf{eq\_refl } x
\end{align*}

@@

We can iteratively apply the above correspondence to obtain an indexed representation of semi-cubical sets.

\begin{align*}
  X_0 & :                  & \U                                                       \\
  X_1 & :                  & X_0 \times X_0 \rightarrow \U                            \\
  X_2 & : \forall a b c d, & X_1 (a, b) \rightarrow X_1 (a, c) \rightarrow X_1 (b, d) \\
      &                    & \rightarrow X_1 (c, d) \rightarrow \U                    \\
  \ldots
\end{align*}

>>>>

% The unit and bool types
\newcommand{\unittype}{\ensuremath{\mathsf{unit}}}
\newcommand{\booltype}{\ensuremath{\mathsf{bool}}}
\newcommand{\unitpoint}{\ensuremath{\ast}}

@@

The outline of the paper is as follows. First, we recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets. We then proceed to explain our indexed construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning with sets inside type theory. We end with some of the finer details of our mechanization.

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner by taking sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$. That is:

@@

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a\emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Coq~\cite{coq22} or Lean~\cite{lean15}. Type theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path and substitutivity of equality as transport. This model supports the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, and the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs. In type theory, logical propositions are types themselves and we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

@@

\begin{align*}
  \Dom & : \Type_m                                        \\
  \UIP & : \Pi x y : \Dom, \Pi p q : x = y, p = q
\end{align*}

@@

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:
    \begin{align*}
      \textsf{eq\_refl}        & : \Pi x, x = x                                                    \\
      \textsf{UIP\_refl\_unit} & : \Pi u : \textsf{unit}, \Pi x : u = u, x = \textsf{eq\_refl} \\
      \textsf{UIP\_refl\_bool} & : \Pi b : \textsf{bool}, \Pi x : b = b, x = \textsf{eq\_refl}
    \end{align*}
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

\begin{align*}
  \textsf{UIP\_refl} & : \Pi x:\Dom, \Pi p : x = x, p = \textsf{eq\_refl } x
\end{align*}

@@

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \times X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions. Based on these coherence conditions, the type of $X_2$ happens to have the form $\Pi a b c d, X_1 (a, b) \times X_1 (a, c) \times X_1 (b, d) \times X_1 (c, d) \rightarrow \U$, recovering the square structure of $X_2$. In summary,

\begin{align*}
  X_0 & :                  & \U                                                       \\
  X_1 & :                  & X_0 \times X_0 \rightarrow \U                            \\
  X_2 & : \Pi a b c d, & X_1 (a, b) \times X_1 (a, c)  \times X_1 (b, d) \times X_1 (c, d) \\
 & &  \rightarrow \U                    \\
  \ldots
\end{align*}
